---
title: "CUT&RUN downstream analysis"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
params:
  config_path: null
  portfolio_snapshot: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 8, fig.height = 5.5)
source(file.path("R", "helpers.R"))
resolve_project_path <- function(path) if (grepl("^/", path)) path else file.path(getwd(), path)
config <- load_analysis_config(params$config_path)
plot_label <- if (!is.null(config$figure_label) && nzchar(config$figure_label)) config$figure_label else config$data_label
paths <- prepare_output_directories(resolve_project_path(config$output_dir))
warning_log <- character()
capture_warnings <- function(expr) withCallingHandlers(expr, warning = function(w) {
  warning_log <<- c(warning_log, conditionMessage(w)); invokeRestart("muffleWarning")
})

counts_path <- resolve_project_path(config$counts_path)
metadata_path <- resolve_project_path(config$metadata_path)
contrasts_path <- resolve_project_path(config$contrasts_path)
sample_metadata <- read_sample_metadata(metadata_path)
inputs <- read_peak_counts(counts_path, config$coordinate_system, sample_ids = rownames(sample_metadata))
contrasts <- read_contrasts(contrasts_path)
sample_metadata <- validate_analysis_inputs(inputs$counts, sample_metadata, contrasts, config$design)

peak_reference <- inputs$peaks
annotation_coverage <- NULL
if (isTRUE(config$annotation$enabled)) {
  peak_reference <- capture_warnings(annotate_peaks_optional(inputs$peaks, config))
  write.csv(peak_reference, file.path(paths[["tables"]], "peak_annotation.csv"), row.names = FALSE)
  annotation_coverage <- data.frame(
    input_peaks = nrow(peak_reference),
    annotated_category = sum(!is.na(peak_reference$annotation) & nzchar(peak_reference$annotation)),
    gene_id_assigned = sum(!is.na(peak_reference$geneId) & nzchar(as.character(peak_reference$geneId))),
    symbol_assigned = sum(!is.na(peak_reference$SYMBOL) & nzchar(as.character(peak_reference$SYMBOL))),
    distance_to_tss_available = sum(is.finite(peak_reference$distanceToTSS)),
    stringsAsFactors = FALSE
  )
  write.csv(annotation_coverage, file.path(paths[["tables"]], "annotation_coverage.csv"), row.names = FALSE)
}

dds <- prepare_deseq_dataset(
  inputs$counts, sample_metadata, config$design, config$min_total_count,
  config$normalization, project_root = getwd()
)
dds <- capture_warnings(fit_deseq_model(dds))
transformed <- capture_warnings(transform_for_visualization(dds))

normalized <- as.data.frame(DESeq2::counts(dds, normalized = TRUE), check.names = FALSE)
normalized$peak_id <- rownames(normalized)
normalized <- merge(inputs$peaks, normalized, by = "peak_id", sort = FALSE)
normalized <- normalized[match(inputs$peaks$peak_id, normalized$peak_id), ]
write.csv(normalized, file.path(paths[["tables"]], "normalized_peak_counts.csv"), row.names = FALSE)
size_factor_table <- data.frame(sample_id = colnames(dds), deseq2_size_factor = DESeq2::sizeFactors(dds))
write.csv(size_factor_table, file.path(paths[["tables"]], "size_factors.csv"), row.names = FALSE)

result_list <- list()
summary_rows <- vector("list", nrow(contrasts))
for (index in seq_len(nrow(contrasts))) {
  contrast <- contrasts[index, , drop = FALSE]
  label <- contrast$label[[1L]]
  ledger <- capture_warnings(run_deseq_contrast(dds, contrast, config$fdr, input_counts = inputs$counts))
  summary_rows[[index]] <- summarize_contrast_result(ledger, label)
  result <- merge(peak_reference, ledger, by = "peak_id", sort = FALSE)
  result <- result[match(inputs$peaks$peak_id, result$peak_id), ]
  result_list[[label]] <- result
  write.csv(result, file.path(paths[["tables"]], paste0(safe_file_stem(label), "_deseq2.csv")), row.names = FALSE)
}
result_summary <- do.call(rbind, summary_rows)
write.csv(result_summary, file.path(paths[["tables"]], "contrast_summary.csv"), row.names = FALSE)

input_files <- c(counts = counts_path, metadata = metadata_path, contrasts = contrasts_path,
                 config = params$config_path)
if (identical(config$normalization$method, "provided_deseq2_size_factors")) {
  input_files <- c(input_files, size_factors = resolve_project_path(config$normalization$size_factors_path))
}
write.csv(input_manifest(input_files), file.path(paths[["tables"]], "input_manifest.csv"), row.names = FALSE)
write.csv(software_versions(unique(c(required_core_packages(),
  if (isTRUE(config$annotation$enabled)) required_annotation_packages(config) else character()))),
  file.path(paths[["tables"]], "software_versions.csv"), row.names = FALSE)
write.csv(code_manifest(c("R/helpers.R", "report/cutandrun_downstream.Rmd", "run_analysis.R")),
          file.path(paths[["tables"]], "code_manifest.csv"), row.names = FALSE)
writeLines(capture.output(dput(config)), file.path(paths[["tables"]], "resolved_config.txt"))
run_metadata <- data.frame(
  field = c("project", "data_status", "genome_build", "input_coordinate_system",
            "analysis_coordinate_system", "design", "normalization", "minimum_total_count",
            "fdr", "input_peaks", "retained_peaks", "samples", "annotation_enabled", "generated_utc"),
  value = c(config$project_title, config$data_label, config$genome_build, config$coordinate_system,
            "one_based_closed", config$design, config$normalization$method, config$min_total_count,
            config$fdr, nrow(inputs$counts), nrow(dds), ncol(inputs$counts),
            isTRUE(config$annotation$enabled), format(Sys.time(), tz = "UTC", usetz = TRUE)),
  stringsAsFactors = FALSE
)
if (!is.null(config$provenance)) {
  run_metadata <- rbind(run_metadata, data.frame(
    field = paste0("provenance_", names(config$provenance)),
    value = unlist(config$provenance, use.names = FALSE), stringsAsFactors = FALSE
  ))
}
write.csv(run_metadata, file.path(paths[["tables"]], "run_metadata.csv"), row.names = FALSE)
```

# Analysis summary

**`r config$project_title`**  
**Data status:** `r config$data_label`

This report starts from an explicit consensus-peak-by-sample integer count matrix. It does not call peaks, count BAM files, reproduce alignment QC, or run DiffBind.

```{r headline-summary}
knitr::kable(result_summary, digits = 4, caption = "Contrast-level result ledger summary")
```

Every input region remains in each result export. Missing adjusted p-values are not called non-significant: all_zero, low_count_filtered, unavailable, and independent_filtered are explicit. Benjamini-Hochberg adjustment and DESeq2 independent filtering are performed separately for each contrast.

# Inputs, model, and normalization boundary

```{r inputs}
knitr::kable(run_metadata, caption = "Resolved run metadata")
knitr::kable(sample_metadata, caption = "Sample metadata")
knitr::kable(contrasts, caption = "Declared comparisons")
knitr::kable(size_factor_table, digits = 4, caption = "Size factors used in the fitted model")
```

Only the coordinate columns plus sample IDs declared in metadata are accepted from the count matrix; unexpected columns cause an error rather than silently becoming count columns. Root-relative paths are resolved from the repository root. Numeric-only categorical codes are rejected.

```{r normalization-note, results='asis'}
if (identical(config$normalization$method, "deseq2_median_ratio")) {
  cat("DESeq2 median-of-ratios normalization estimates **relative** occupancy and assumes most retained peaks do not shift together. Without spike-in or another external scale, this analysis cannot establish a global gain or loss of signal.\n")
} else {
  cat("Externally supplied DESeq2 size-factor divisors are used. Their experimental validity remains the analyst's responsibility.\n")
}
```

# Sample diagnostics

```{r diagnostics}
library_qc <- make_library_qc_plot(inputs$counts, sample_metadata,
  paste(plot_label, "library depth and detected peaks", sep = " - "))
save_ggplot(library_qc, file.path(paths[["figures"]], "library_qc.png"))
library_qc
pca_plot <- make_pca_plot(transformed, sample_metadata, config$pca_top_peaks,
  paste(plot_label, "variance-stabilized peak-count PCA", sep = " - "))
save_ggplot(pca_plot, file.path(paths[["figures"]], "pca.png"))
pca_plot
correlation_path <- file.path(paths[["figures"]], "sample_correlation.png")
correlation <- save_correlation_heatmap(transformed, correlation_path,
  paste(plot_label, "sample correlation", sep = " - "), metadata = sample_metadata)
write.csv(correlation, file.path(paths[["tables"]], "sample_correlation.csv"))
knitr::include_graphics(normalizePath(correlation_path))
dispersion_path <- file.path(paths[["figures"]], "dispersion_estimates.png")
save_dispersion_plot(dds, dispersion_path, paste(plot_label, "dispersion estimates", sep = " - "))
knitr::include_graphics(normalizePath(dispersion_path))
```

No samples are excluded automatically from these diagnostics. In the public example each condition has only two biological replicates; the separated untreated pair is retained as observed. With n=2 per condition, sample-level influence judgments are fragile, and standard Cook's-distance replacement is not available as it is for larger replicate groups.

# Differential peaks

Heatmaps select the lowest adjusted p-values, then largest absolute fold changes, then peak ID. They display row-centered VST counts for the selected significant regions; samples remain in input order.

```{r differential, results='asis'}
for (label in names(result_list)) {
  result <- result_list[[label]]
  stem <- safe_file_stem(label)
  cat("\n## ", label, "\n\n", sep = "")
  print(knitr::kable(result_summary[result_summary$contrast == label, ], digits = 4))
  ma_plot <- make_ma_plot(result, label, config$fdr, plot_label)
  save_ggplot(ma_plot, file.path(paths[["figures"]], paste0(stem, "_ma.png")))
  print(ma_plot)
  volcano <- make_volcano_plot(result, label, config$fdr, plot_label)
  save_ggplot(volcano, file.path(paths[["figures"]], paste0(stem, "_volcano.png")))
  print(volcano)
  heatmap_path <- file.path(paths[["figures"]], paste0(stem, "_significant_peak_heatmap.png"))
  heatmap_info <- save_significant_peak_heatmap(
    transformed, sample_metadata, heatmap_path, result, config$fdr, config$heatmap_top_peaks,
    paste(plot_label, label, "significant peak replicate heatmap", sep = " - ")
  )
  cat("\nThe heatmap shows ", heatmap_info$selected_n, " of ",
      heatmap_info$significant_n, " significant regions.\n\n", sep = "")
  print(knitr::include_graphics(normalizePath(heatmap_path)))
}
```

# Genomic annotation

```{r annotation, results='asis'}
if (isTRUE(config$annotation$enabled)) {
  print(knitr::kable(annotation_coverage, caption = "Annotation coverage"))
  cat("\nAnnotation used ", config$annotation$txdb_package, " and ",
      config$annotation$orgdb_package, ". Genome build and seqnames were validated. ",
      "Each contrast export includes annotation category, nearest gene fields, and distanceToTSS.\n", sep = "")
} else {
  cat("Annotation is disabled. This prevents synthetic coordinates from being presented as a reference genome.\n")
}
```

# Interpretation boundary

The exported tables support review of relative differential peaks, annotation, and replicate patterns. The workflow intentionally omits exploratory derivative networks, online enrichment, WikiPathways, Cytoscape examples, cross-mark analyses, and repeat-family/LTR10 expansion; each would need a separately declared design and reference provenance.

# Downloads and reproducibility

```{r downloads, results='asis'}
warning_table <- data.frame(warning = unique(warning_log), stringsAsFactors = FALSE)
if (!nrow(warning_table)) warning_table <- data.frame(warning = "No warnings captured.")
write.csv(warning_table, file.path(paths[["tables"]], "warnings.csv"), row.names = FALSE)
writeLines(capture.output(sessionInfo()), file.path(paths[["tables"]], "session_info.txt"))
if (isTRUE(params$portfolio_snapshot)) {
  cat("This saved portfolio snapshot links only to compact tracked results. Full tables are regenerated under the configured output directory.\n\n")
  snapshot_files <- c(
    "Contrast summary" = "../results/gse186608/contrast_summary.csv",
    "Top significant-region snapshot" = "../results/gse186608/top_significant_regions.csv"
  )
  cat(paste0("- [", names(snapshot_files), "](", snapshot_files, ")", collapse = "\n"))
} else {
  table_files <- sort(list.files(paths[["tables"]]))
  cat(paste0("- [", table_files, "](tables/", table_files, ")", collapse = "\n"))
}
```

The manifests record resolved input paths and SHA-256 hashes, code hashes, resolved configuration, software versions, warnings, and the full R session. They document the executed environment without claiming an automatically restorable lockfile.

```{r session-info}
sessionInfo()
```
